Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2017-03-08 manuscript/manuscript.Rnw
move the legend in the CPT shape plot
don't use reactome; discuss changes in GSEA due to core change

<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
pathway <- list()
for (database in c("kegg","reactome","biocarta","netpath")) {
    load(paste0("../data/placenta_gsea_",database))
    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
    pathway[[database]] <-
        gsea.table
}
pathway.analysis <-
    rbindlist(pathway)[fdr < 0.1,]
rm(pathway)
pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                gsub("_+"," ",
                                     Pathway))]
pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]

@@

      + geom_segment(mapping=aes(y=V1-ifelse(V1<80,20,0),

@@

                                                             max(`Expression (FPKM)`)*4]

@@

                                                                  max(`Expression (FPKM)`)*4.1]

@@

                      y=V1-ifelse(V1<80,20,0),
                      label=paste0(V1," genes"),override=TRUE),

@@

                         y=100,

@@

                      y=100,
                      label=paste0(V1," genes"),
                      override=TRUE),

@@

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome %
%
(\Sexpr{placpath("^RIBOSOME","Kegg")})
, translation %
(\Sexpr{placpath("^TRANSLATION","Reactome")})
%
, signal
recognition particle targeting to the rough endoplasmic reticulum %
(\Sexpr{placpath("^SRP DEPENDENT","Reactome")})
%
,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway %
(\Sexpr{placpath("^EGFR1","Netpath")})
%
and the Actin Y pathway %
(\Sexpr{placpath("^ACTINY","Biocarta")})
%
. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \ac{ITGB1} and its binding
protein \ac{ITGB1BP1} as well as \ac{FN1} are highly expressed, which
activates Rac. This potentially leads to the formation of focal adhesions
and lamellipodia formation through inhibition of GSN and
HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
which forms tight junctions in epithelia, the proteoglycans \ac{SDC4} and
\ac{LUM}, and \ac{LAMC1}.

>>>>

<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
pathway <- list()
for (database in c("kegg","biocarta","netpath")) {
    load(paste0("../data/placenta_gsea_",database))
    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
    pathway[[database]] <-
        gsea.table
}
pathway.analysis <-
    rbindlist(pathway)[fdr < 0.1,]
rm(pathway)
pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                gsub("_+"," ",
                                     Pathway))]
pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]

@@

      + geom_segment(mapping=aes(y=V1-ifelse(V1<100,20,0),

@@

                                                             max(`Expression (FPKM)`)*2]

@@

                                                                  max(`Expression (FPKM)`)*2.1]

@@

                      y=V1-ifelse(V1<100,20,0),
                      label=paste0(V1," genes")),inherit.aes=FALSE,

@@

                         y=10^2.5,

@@

                      y=10^2.5,
                      label=paste0(V1," genes")),
                      inherit.aes=FALSE,

@@

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways (Supplemental
\cref{tab:core_placenta_pathway_analysis}). The most abundant set
of significantly enriched molecular pathways are related to the
migration of the trophoblast into the endometrium, including the EGFR1
pathway %
(\Sexpr{placpath("^EGFR1","Netpath")}), Focal Adhesion pathway
(\Sexpr{placpath("^EGFR1","Kegg")})
%
and the Actin Y pathway %
(\Sexpr{placpath("^ACTINY","Biocarta")})%
%
. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \ac{ITGB1} and its binding
protein \ac{ITGB1BP1} as well as \ac{FN1} are highly expressed, which
activates Rac. This potentially leads to the formation of focal adhesions
and lamellipodia formation through inhibition of GSN and
HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
which forms tight junctions in epithelia, the proteoglycans \ac{SDC4} and
\ac{LUM}, and \ac{LAMC1}.

### 2017-03-08 manuscript/supplemental_material.Rnw
move the legend in the supplemental CPTL shape plot
don't use reactome; discuss changes in GSEA due to core change

      + geom_segment(mapping=aes(y=V1-ifelse(V1<145,40,0),

@@

                      y=V1-ifelse(V1<145,40,0),
                      label=paste0(V1," genes"),override=TRUE),

@@

              legend.position=c(0.99,0.99),

@@

              legend.justification=c(1,1),

@@

\begin{table}
<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
pathway <- list()
for (database in c("kegg","reactome","biocarta","netpath")) {
    load(paste0("../data/placenta_gsea_",database))
    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
    pathway[[database]] <-
        gsea.table
}
pathway.analysis <-
    rbindlist(pathway)[fdr < 0.1,]
rm(pathway)
pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                gsub("_+"," ",
                                     Pathway))]
pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]

setnames(pathway.analysis,"p_values","p")
setorder(pathway.analysis,"p")
print(xtable(pathway.analysis,
             display=c("s","s","g","g","s"),
             align=c("c","p{3.5in}","r","r","c")),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE)      
@ 
\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
  biocarta, reactome, and kegg pathway databases}
\label{tab:core_placenta_pathway_analysis}
\end{table}

>>>>

      + geom_segment(mapping=aes(y=V1-ifelse(V1<160,40,0),

@@

                      y=V1-ifelse(V1<160,40,0),
                      label=paste0(V1," genes")),
                  inherit.aes=FALSE,

@@

              legend.position=c(0.01,0.01),

@@

              legend.justification=c(0,0),

@@

\begin{table}
<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
pathway <- list()
for (database in c("kegg","biocarta","netpath")) {
    load(paste0("../data/placenta_gsea_",database))
    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
    pathway[[database]] <-
        gsea.table
}
pathway.analysis <-
    rbindlist(pathway)[fdr < 0.1,]
rm(pathway)
pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                gsub("_+"," ",
                                     Pathway))]
pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]

setnames(pathway.analysis,"p_values","p")
setorder(pathway.analysis,"p")
print(xtable(pathway.analysis,
             display=c("s","s","g","g","s"),
             align=c("c","p{3.5in}","r","r","c")),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE)      
@ 
\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
  biocarta, netpath, and kegg pathway databases}
\label{tab:core_placenta_pathway_analysis}
\end{table}
